\bf Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-17 appendix.tex
Updated the appendix
Update Figure 2 of appendix section 2
Added scDEED section

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

>>>>

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand{\To}{\textbf{to}~}
\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\subsection{Single-cell gene
expression}\label{single-cell-gene-expression}

\subsubsection{Comparison with results of scDEED
recommendations}\label{comparison-with-results-of-scdeed-recommendations}

As we were writing this paper \citet{xia2023} appeared proposing a new
method called scDEED helping to assess the validity of a \gD{}
embedding. scDEED calculates a reliability score for each cell embedding
based on the similarity between the cell's \gD{} embedding neighbors and
its neighbors prior to embedding. A low reliability score suggests a
dubious embedding. It can help in the deciding on optimal
hyper-parameters. Here we illustrate how our method compares with the
results from scDEED

Following the process in \citet{xia2023} again using the Human
Peripheral Blood Mononuclear Cells (PBMC) data. Note that
\citet{xia2023} uses a different subset from \textbf{?@sec-pbmc}, which
contains \(31,021\) cells including cell type labels, and the gene
expression levels were in the unit of log-transformed UMI count per
\(10,000\). They focused on three sequencing methods (inDrops, DropSeq,
and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell,
CD14+ Monocyte, and B cell.

For illustration purposes, we only selected cells generated with inDrops
(\(n=5858\) cells) and UMAP cell embeddings. Also, \citet{xia2023} used
first \(50\) principal components to generate the UMAP. The objective is
to assess the optimized layout by scDEED, and if it does not accurately
represent the three clusters with small separations of the PBMC dataset,
then select a reasonable \gD{} layout.

\textbf{?@fig-model-pbmc-author-proj-scdeed} shows layout from the
original published article along with that suggested by scDEED to be
more accurate. The log MSE vs binwidth plot illustrates that our
approach would suggest that scDEED is correct here, that layout b is
more accurately reflecting the cluster structure in the data.

\includegraphics[width=1\textwidth,height=\textheight]{appendix_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

\caption{\label{fig-pbmc-mse-umap}Assessing which of the 2 UMAP layouts
with different hyperparameter settings (n\_neighbors: \(30\), min\_dist:
\(0.3\) (red), n\_neighbors: \(80\), min\_dist: \(0.5\) (blue)) on the
PBMC data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the left plot and in
the scatterplots represents UMAP layout (a, b). Of the two, layout b is
optimal across all binwidths making it the best choice.}
